---
Title: "P4 White Wine Project"
Author: "Shannon Brinkmeyer"
Date: "5/22/2017"
output: html_document
---
# This report explores in depth numerous different White Wines.
  
  
# Load the Packages
``` {r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Packages}

library(dplyr)
library(GGally)
library(ggplot2)
library(gridExtra)
library(memisc)
library(car)
library(reshape2)
library(dplyr)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

WW <- read.csv('~/Desktop/P4/wineQualityWhites.csv')

```

# Univariate Plots Section

``` {r echo=FALSE, message=FALSE, warning=FALSE, Explore_Data}

names(WW)

summary(WW)

```

### This dataset includes 11 different input variables with over 4898 observations 
### of White Wines and 1 output variable (Quality).

```{r echo=FALSE, message=FALSE, warning=FALSE, Remove_Variable}

WW$X <- NULL

```

### Remove the unneeded variable X.


```{r echo=FALSE, message=FALSE, warning=FALSE, Histograms}

hist(WW$alcohol)

hist(WW$citric.acid)

```

### 2 quick histogram charts to show the frequency of alcohol and citric acid in
### the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$fixed.acidity, binwidth=.1)

qplot(WW$volatile.acidity, binwidth=.01)

qplot(WW$density, binwidth=.0001)

qplot(WW$sulphates, binwidth=.01)

```

### These plots show a mostly normal distribution

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$chlorides, binwidth=.001)

qplot(WW$pH, binwidth=.01)

qplot(WW$quality, binwidth=.5)

qplot(WW$free.sulfur.dioxide, binwidth=2)

qplot(WW$total.sulfur.dioxide, binwidth=5)

```

### These plots also show a mostly normal distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$residual.sugar, binwidth=.5)

qplot(WW$alcohol, binwidth=.1)

```

### The residual.sugar plot is skewed to the left (less sweet White Wines?) 
### and the alcohol plot is pretty spread out.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$citric.acid, binwidth=.01)

```

### There is an interesting spike in citric acid around .5.

# Univariate Analysis

## What is the structure of your dataset?

### The Dataset is made up of 4898 observation of White Wines with 11 inputs 
### (fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, 
### free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol) 
### and 1 output (quality).

## What is the main features of interest in your dataset?

### I think the main features of this White Wine dataset are Alcohol(%) as
### well as Residual Sugar. They are the 2 main variables that appear to not
### have a normal distribution.

## What other features in the dataset do you think will help support your
## investigation into your feature(s) of interest?

### Chlorides, Volatile Acidity, and total sulfur dioxide seem to play a smaller 
### part in the quality of the White Wine. Citric acid also has an interesting
### spike around .5.

## Did you create any new variables from existing variables in the dataset?

### I created a new variable called quality_fac to aid in the factoring of the 
### quality of some of my plots and to show better visualizations of the data
### in the Multivariate Section below.

## Of the features you investigated, were there any unusual distributions? 
## Did you perform any operations on the data to tidy, adjust, or change the 
## form of the data? If so, why did you do this?

### The only unusual distribution I found was sulphates which seemed to have a 
### minor bimodal distribution. They only change I made was the new variable as  
### mentioned above.


# Biavariate Plots

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations}

ggpairs(WW, 1:12,
        lower = list(continuous = wrap('points', shape = I('.'))),
        upper = list(combo = wrap('box', outlier.shape = I('.'))))

```

### Using GGpairs to see any apparent correlations with the data. There appears 
### to be multiple correlations between a number of variables in the dataset to 
### explore including density/residual sugar and alcohol/quality, among others.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plots}

ggplot(aes(x = quality, y = pH), data = WW) +
  geom_point()

ggplot(aes(x = quality, y = alcohol), data = WW) +
  geom_point()

ggplot(aes(x = quality, y = citric.acid), data = WW) +
  geom_point()

ggplot(aes(x = quality, y = residual.sugar), data = WW) +
  geom_point()

```

### The best quality White Wines seem to have a pH of 3.0 to 3.5, alcohol content
### of between 10 and 13, medium to high levels of citric acid (.25-.5), and low  
### residual sugars(0 - 18).

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = quality, y = density), data = WW) +
  geom_point()

ggplot(aes(x = quality, y = chlorides), data = WW) +
  geom_point()

ggplot(aes(x = quality, y = sulphates), data = WW) +
  geom_point()

ggplot(aes(x = quality, y = fixed.acidity), data = WW) +
  geom_point()

```

### They also seen to have a lower density, lower chlorides (.2 - .6), somewhat
### lower amount of sulphates, and fixed acidity between 4 - 8.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(fixed.acidity, residual.sugar, data=WW) +
  geom_point(color='red') +
  scale_x_continuous(lim=c(unname(quantile(WW$fixed.acidity, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.99))))) +
  geom_smooth(method='lm')

qplot(density, total.sulfur.dioxide, data=WW) +
  geom_point(color='red') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$total.sulfur.dioxide, 
                                           c(0.01,0.99))))) +
  geom_smooth(method='lm')

qplot(density, alcohol, data=WW) +
  geom_point(color='red') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.99))))) +
  geom_smooth(method='lm')

qplot(density, residual.sugar, data=WW) +
  geom_point(color='red') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.99))))) +
  geom_smooth(method='lm')

```

### There is definitely a linear relationship between density and total sulfur
### dioxide as seen in the plot above as well as a negative linear relationship 
### between alcohol and density. After comparing density and residual sugar, 
### they appears to be a very strong linear relationship also.


# Bivariate Analysis

## Talk about some of the relationships you observed in this part of the 
## investigation. How did the feature(s) of interest vary with other features in 
## the dataset?

### The GGpairs plot was pretty interesting because it put everything together 
### and showed correlations between the variables. One of the biggest was the 
### correlation between density and residual sugars as well as density and 
### alcohol. 
### All the higher quality White Wines have a medium level of pH between 3-3.5,
### higher level of alcohol content, mid-high citric acid level, lower residual
### sugars, lower chlorides, lower density, and somewhat lower fixed acidity.

## Did you observe any interesting relationships between the other features
## (not the main feature(s) of interest)?

### Density seems to be closely related to residual sugar and fixed acidity. 
### Alcohol also seems to be related to density and total sulfur dioxide, and it
### appears the higher alcohol content it has the less dense the White Wine is.

## What was the strongest relationship you found?

### The strongest relationship appears to be between Density and Residual Sugar.


# Mulivariate Plot Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

WW$quality_fac <- as.factor(WW$quality)

```

### Adding factor to quality for ranking purposes.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density, y = alcohol, colour = quality_fac), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95))))) +
  scale_color_brewer(type = 'div', palette = 4)

```

### The higher quality White Wines seem to have more alcohol and lower density
### than the lower quality White Wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(sulphates, total.sulfur.dioxide, data=WW, color=quality_fac, geom='point')

qplot(sulphates, free.sulfur.dioxide, data=WW, color=quality_fac, geom='point')

```

### Both plots show no real correlation between sulphates and total or free 
### sulfur dioxide on the quality of White Wine. I was curious because sulphates
### tend to contribute to sulfur dioxide levels according to the description of
### attributes.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(fixed.acidity, volatile.acidity, data=WW, color=quality_fac) +
  geom_point(alpha = 1, size = 1, position = 'jitter') +
  geom_smooth(method='lm') 

```

### Volatile Acidity seems to play a role in the quality of White Wine as well
### as Fixed Acidity to a lower extent.

```{r echo=FALSE, message=FALSE, warning=FALSE}

p1 <- ggplot(aes(citric.acid, chlorides, colour=quality_fac), data=WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$citric.acid, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$chlorides, c(0.01,0.95)))))

p2 <- ggplot(aes(density, residual.sugar, colour=quality_fac), data=WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.95)))))

p3 <- ggplot(aes(residual.sugar, free.sulfur.dioxide, colour=quality_fac), 
             data=WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$free.sulfur.dioxide, 
                                           c(0.01,0.95)))))

p4 <- ggplot(aes(alcohol, density, colour=quality_fac), data=WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95)))))

p5 <- ggplot(aes(alcohol, total.sulfur.dioxide, colour=quality_fac), data=WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$total.sulfur.dioxide, 
                                           c(0.01,0.95)))))

p6 <- ggplot(aes(fixed.acidity, citric.acid, colour=quality_fac), data=WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$fixed.acidity, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$citric.acid, c(0.01,0.95)))))

grid.arrange(p1, p2, p3, p4, p5, p6, ncol=2)

```

### Quality White Wines have an above average level of citric acid, lower level of 
### chlorides, higher level of alcohol, and medium to high level of fixed acidity
### compared to the lower quality White Wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, alcohol, colour=quality_fac), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95)))))

```

### Once again the Quality of White Wines but this time with the Density and 
### Alcohol switched around showing the strong linear relationship as well as the
### quality factor.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density, y = alcohol, color = quality_fac), data = WW) +
  geom_point(aes()) +
  scale_color_brewer(type = 'qual', palette = 2)

ggplot(aes(x = density, y = alcohol, colour = quality_fac), data = WW) +
  facet_wrap( ~ quality) +
  geom_point(aes()) +
  scale_color_brewer(type = 'qual', palette = 2)
```

### This is a better look at the different quality White Wines compared to   
### Density and Alcohol together and then seperate in order to see the 
### difference.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(alcohol, citric.acid, data= WW, color = quality_fac, geom='boxplot')

qplot(alcohol, density, data= WW, color = quality_fac, geom='boxplot')

qplot(alcohol, chlorides, data= WW, color = quality_fac, geom='boxplot')

```

### All three of these boxplots seems to back up my findings that higher citric
### acid, lower density (higher alcohol content), and lower chlorides make a 
### better quality White Wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, total.sulfur.dioxide, colour=alcohol), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$total.sulfur.dioxide, 
                                           c(0.01,0.95)))))

```

### Density seems to be closely tied to the alcohol content as well as possibly
### total sulfur dioxide. The higher the alcohol content the less dense the 
### White Wines appear to be.


# Multivariate Analysis

## Talk about some of the relationships you observed in this part of the 
## investigation. Were there features that strengthened each other in terms of 
## looking at your feature(s) of interest?

### I was suprised to find that the higher quality White Wines seem to have a 
### higher alcohol content which in turn means a lower density. I thought that  
### the opposite would be true due to the taste of alcohol.
### I was also suprised to find that the higher quality White Wines had a medium 
### to high level of citric acid as well as low levels of chlorides(salt).

## Were there any interesting or surprising interactions between features?

### I thought it was definitely interesting that as the alcohol content goes 
### up the density goes down.

## OPTIONAL: Did you create any models with your dataset? Discuss the
## strengths and limitations of your model.

### I did not create a model.


# Final Plots and Summary

# Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

qplot(density, residual.sugar, data=WW) +
  geom_point(color='red') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.99))))) +
  geom_smooth(method='lm') +
  labs(x='Density', y='Residual Sugar', title='Strong Linear Relationship')
  
```

### The Density and Residual Sugar of the White Wines have a strong linear 
### relationship as shown in the above plot

# Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

ggplot(aes(x = density, y = alcohol, colour = quality_fac), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.99))))) +
  scale_color_brewer(type = 'div', palette = 4) +
  labs(x='Density', y='Alcohol', colour='Quality', 
       title='The Makeup of a Quality White Wine')

```

### The higher quality White Wines have a higher alcohol content and lower 
### density than the lower quality White Wines. It also shows a strong linear
### relationship between density and alcohol.

# Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

qplot(alcohol, citric.acid, data= WW, color = quality_fac, geom='boxplot') +
  labs(x='Alcohol (%)', y='Citric Acid (g/dm^3)', title='Higher Citric Acid 
       Makes a Higher Quality White Wine', color='Quality')

qplot(alcohol, density, data= WW, color = quality_fac, geom='boxplot') +
  labs(x='Alcohol (%)', y='Density (g/dm^3)', color='Quality', title='Lower 
       Density Makes a Higher Quality White Wine')

qplot(alcohol, chlorides, data= WW, color = quality_fac, geom='boxplot') +
  labs(x='Alcohol (%)', y='Chlorides (g/dm^3)', color='Quality', title='Lower 
       Chlorides Makes a Higher Quality White Wine')

```

### All three of these boxplots seems to back up my findings that higher citric
### acid, lower density (higher alcohol content), and lower chlorides make a 
### better quality White Wine.


# Reflection

### This dataset contained 4898 observations of White Wine Quality with 11
### inputs and 1 output. After exploring the data in detail I can say for certain 
### I know alot more about Wine than I have ever known. At first I was 
### concentrating strictly on what variables are needed to make a high quality
### wine then after more research I started wondering how the variables related
### with one another. I was very suprised to find that density is closely related
### to alcohol content as well as residual sugars. The more dense the wine was 
### the less alcohol content it contained.

### After examing the sulphates and sulfur dioxide I was very suprised to learn
### they are not closely correlated as it mentioned in the description of 
### attributes that sulphates can contribute to sulfur dioxide gas levels. It
### appears that density and total sulfur dioxide have a linear realtionship also
### that could be futhur examined.

### I think there is opportunity for furthur understanding of the makeup of
### a good quality wine with more data on a wider number of white wines. Breaking 
### up the data into the 7 classes of whites would also allow you to gain more 
### understanding how the different variables that make up the wines react and
### come together to form a high quality wine. 








