---
Title: "P4 White Wine Project"
Author: "Shannon Brinkmeyer"
Date: "5/22/2017"
output: html_document
---
# This report explores in depth numerous different White Wines.
  
# Load the Packages
``` {r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Packages}

library(dplyr)
library(GGally)
library(ggplot2)
library(gridExtra)
library(memisc)
library(car)
library(reshape2)
library(dplyr)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

WW <- read.csv('~/Desktop/P4/wineQualityWhites.csv')

```


# Univariate Plots Section

``` {r echo=FALSE, message=FALSE, warning=FALSE, Explore_Data}

names(WW)

summary(WW)

```

### This dataset includes 11 different input variables with over 4898  
### observations of White Wines and 1 output variable (Quality).

```{r echo=FALSE, message=FALSE, warning=FALSE, Remove_Variable}

WW$X <- NULL

```

### I have decided to remove the uneeded variable X because it surves no real 
### purpose but to be used as a counter.



```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$fixed.acidity, binwidth=.1)

qplot(WW$volatile.acidity, binwidth=.01)

qplot(WW$density, binwidth=.0001)

qplot(WW$sulphates, binwidth=.01)

```

### I decided to do a quick plot with every variable to try and better
### understand the data and also to see the distribution. These plots are
### of a normal distribution type. There appears to be numerous outliers with 
### pretty high counts of fixed acidity, volatile acidity, and sulphates.
### Density appears to be the only plot with a limited number of outliers.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$chlorides, binwidth=.001)

summary(WW$chlorides)

qplot(WW$pH, binwidth=.01)

summary(WW$pH)

qplot(WW$quality, binwidth=.5)

summary(WW$quality)

qplot(WW$free.sulfur.dioxide, binwidth=2)

summary(WW$free.sulfur.dioxide)

qplot(WW$total.sulfur.dioxide, binwidth=5)

summary(WW$free.sulfur.dioxide)

```

### After plotting the above plots they also show a mostly normal distribution. 
### Clorides seem to be the only real prevailing variable with a decient amount
### of outliers. pH, free and total sulfur also have a couple of outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$residual.sugar, binwidth=.5)

summary(WW$residual.sugar)

qplot(WW$alcohol, binwidth=.1)

summary(WW$alcohol)

```

### After plotting the residual sugar plot it is skewed to the left (prehaps it 
### is showing less White Wines in the distribution?) and the alcohol plot is 
### pretty spread out.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$residual.sugar, binwidth=.05) +
  scale_x_log10()

summary(WW$residual.sugar)

```

### After transforming the data using a log tranformation is shows a bi-modal
### distribution. White Wines are considered sweet if they have a residual sugar 
### content greater then 45, and it appears there is less sweet White Wines with  
### a median and mean of residual sugars of approx. 5.2-6.4 and a Max of 65.8.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(WW$citric.acid, binwidth=.01)

qplot(WW$citric.acid, binwidth=.05) +
  scale_x_log10()

summary(WW$citric.acid)

```

### After looking at the citric acid distribution it appears there is an 
### interesting spike in citric acid around .5. And after transforming the data  
### with log10 it appears to be mostly normal.



# Univariate Analysis

## What is the structure of your dataset?

### The Dataset is made up of 4898 observation of White Wines with 11 inputs 
### (fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, 
### free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol) 
### and 1 output (quality).

## What is the main features of interest in your dataset?

### I think the main features of this White Wine dataset are Alcohol(%) as
### well as Residual Sugar. They are the 2 main variables that appear to not
### have a normal distribution.

## What other features in the dataset do you think will help support 
## your investigation into your feature(s) of interest?

### Chlorides, Volatile Acidity, and total sulfur dioxide seem to play a smaller 
### part in the quality of the White Wine. Citric acid also has an interesting
### spike around .5.

## Did you create any new variables from existing variables in the dataset?

### I created a new variable called quality_fac to aid in the factoring of the 
### quality of some of my plots and to show better visualizations of the data
### in the Multivariate Section below.

## Of the features you investigated, were there any unusual distributions? 
## Did you perform any operations on the data to tidy, adjust, or change the 
## form of the data? If so, why did you do this?

### Sulphates seemed to have a minor bimodal distribution and residual sugar
### had a bimodal distribution when transformed with log10 . They only change   
### I made was the new variable as mentioned above.


# Biavariate Plots

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations}

ggpairs(WW, 1:12,
        lower = list(continuous = wrap('points', shape = I('.'))),
        upper = list(combo = wrap('box', outlier.shape = I('.'), size=6)))
  
```

### Using GGpairs to see any apparent correlations with the data. There appears 
### to be multiple correlations between a number of variables in the dataset to 
### explore including density/residual sugar and alcohol/quality, among many 
### I will explore.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plots}

ggcorr(WW, hjust = .90, size=5, layout.exp =2)

```

### This plot is also a great example of the correlation between all the 
### variables. It shows the strong relations between residual sugar/density and 
### alcohol/density as well as many others to explore.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(factor(quality), y = pH), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$pH)

ggplot(aes(factor(quality), y = alcohol), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$alcohol)

ggplot(aes(factor(quality), y = citric.acid), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$citric.acid)

ggplot(aes(factor(quality), y = residual.sugar), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$residual.sugar)

```

### I am comparing quality to numerous other variables to get a sense as to what
### goes into making a high quality White Wine.
### The best quality White Wines seem to have a pH of 3.0 to 3.5, alcohol content
### of between 10 and 13, medium to high levels of citric acid (.25-.5), and low  
### residual sugars(0 - 18). I was pretty suprised to see the high alcohol count
### as being a pretty good factor in high quality white wines. It also appears
### the lower residual sugars wines are more appealing.

```{r echo=FALSE, message=FALSE, warning=FALSE}

by(WW$alcohol, WW$quality, summary)

by(WW$residual.sugar, WW$quality, summary)

```
### In comparing the summaries of the variables together it appears the numbers
### with the alcohol content show a trend toward higher levels as the quality 
### gets better, but the numbers on the residual sugars are not so apparent.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(factor(quality), y = density), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$density)
  
ggplot(aes(factor(quality), y = chlorides), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$chlorides)

ggplot(aes(factor(quality), y = sulphates), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$sulphates)

ggplot(aes(factor(quality), y = fixed.acidity), data = WW) +
  geom_jitter(alpha=.3) +
  geom_boxplot( alpha = .5,color = 'blue')

cor(WW$quality, WW$fixed.acidity)

```

### They also seen to have a lower density, lower chlorides (.2 - .6), somewhat
### lower amount of sulphates, and fixed acidity between 4 - 8. It appears lower
### density is related to the quality as well as lower chlorides. I figured 
### the opposite would be true with the chlorides and possibly density.

```{r echo=FALSE, message=FALSE, warning=FALSE}

by(WW$density, WW$quality, summary)

by(WW$chlorides, WW$quality, summary)

```

### It appears the higher quality wines do have lower density and chlorides. 
### Apparently the less sweet and salty wines are of better quality as shown
### by the summary comparisons.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(fixed.acidity, residual.sugar, data=WW, color = factor(quality)) +
  geom_point(color='red', alpha=.5) +
  scale_x_continuous(lim=c(unname(quantile(WW$fixed.acidity, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.99))))) +
  geom_smooth(method='lm')

cor(WW$fixed.acidity, WW$residual.sugar)

qplot(density, total.sulfur.dioxide, data=WW, color = factor(quality)) +
  geom_point(color='red', alpha=.5) +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$total.sulfur.dioxide, 
                                           c(0.01,0.99))))) +
  geom_smooth(method='lm')

cor(WW$density, WW$total.sulfur.dioxide)

qplot(density, alcohol, data=WW, color = factor(quality)) + 
  geom_point(color='red', alpha=.5) +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.99))))) +
  geom_smooth(method='lm')

cor(WW$density, WW$alcohol)

qplot(density, residual.sugar, data=WW, color = factor(quality)) +
  geom_point(color='red', alpha=.5) +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.99))))) +
  geom_smooth(method='lm')

cor(WW$density, WW$residual.sugar)

```
### There is definitely a linear relationship between density and total sulfur
### dioxide as seen in the plot (.53) above as well as a negative linear 
### relationship between alcohol and density (-.78). After comparing density and  
### residual sugar, they appears to have a very strong linear relationship also 
### (.839). There doesn't appear to be any relation between fixed acidity and 
### residual sugar. I guess I wasn't suprised by any of the findings except the 
### density and total sulfur dioxide, I thought density was more tied to the 
### sugar/alcohol content.

# Bivariate Analysis

## Talk about some of the relationships you observed in this part of the 
## investigation. How did the feature(s) of interest vary with other features in 
## the dataset?

### The GGpairs plot was pretty interesting because it put everything together 
### and showed correlations between the variables. One of the biggest was the 
### correlation between density and residual sugars as well as density and 
### alcohol. 
### All the higher quality White Wines have a medium level of pH between 3-3.5,
### higher level of alcohol content, mid-high citric acid level, lower residual
### sugars, lower chlorides, lower density, and somewhat lower fixed acidity.

## Did you observe any interesting relationships between the other features
## (not the main feature(s) of interest)?

### Density seems to be closely related to residual sugar and fixed acidity. 
### Alcohol also seems to be related to density and total sulfur dioxide, and it
### appears the higher alcohol content it has the less dense the White Wine is.

## What was the strongest relationship you found?

### The strongest relationship appears to be between Density and Residual Sugar
### (.839) because as sugar builds in the wines so does the density. 


# Mulivariate Plot Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

WW$quality_fac <- as.factor(WW$quality)

```

### Adding factor to quality for ranking purposes.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density, y = alcohol, colour = quality_fac), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95))))) +
  scale_color_brewer(type = 'div', palette = 4)

```

### I wanted to explore the density and alcohol relationship more due to the 
### strength it has with the quality factor of wines. It is pretty apparent
### that the quality of wines goes up with the higher alcohol levels and lower 
### density levels which also directly affect each other.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(sulphates, total.sulfur.dioxide, data=WW, color=quality_fac, 
      geom='point', alpha=.5)

qplot(sulphates, free.sulfur.dioxide, data=WW, color=quality_fac,
      geom='point', alpha=.5)

```

### Both plots show no real correlation between sulphates and total or free 
### sulfur dioxide on the quality of White Wine. I was curious because sulphates
### tend to contribute to sulfur dioxide levels according to the description of
### attributes.


```{r echo=FALSE, message=FALSE, warning=FALSE}

p1 <- ggplot(aes(citric.acid, chlorides, colour=quality_fac), data=WW) +
  geom_point(alpha = .3, size = 1, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$citric.acid, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$chlorides, c(0.01,0.95)))))

p2 <- ggplot(aes(density, residual.sugar, colour=quality_fac), data=WW) +
  geom_point(alpha = .3, size = 1, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.95)))))

p3 <- ggplot(aes(residual.sugar, free.sulfur.dioxide, colour=quality_fac), 
             data=WW) +
  geom_point(alpha = .3, size = 1, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$free.sulfur.dioxide, 
                                           c(0.01,0.95)))))

p4 <- ggplot(aes(alcohol, density, colour=quality_fac), data=WW) +
  geom_point(alpha = .3, size = 1, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95)))))

p5 <- ggplot(aes(alcohol, total.sulfur.dioxide, colour=quality_fac), data=WW) +
  geom_point(alpha = .3, size = 1, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$total.sulfur.dioxide, 
                                           c(0.01,0.95)))))

p6 <- ggplot(aes(fixed.acidity, citric.acid, colour=quality_fac), data=WW) +
  geom_point(alpha = .3, size = 1, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$fixed.acidity, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$citric.acid, c(0.01,0.95)))))

grid.arrange(p1, p2, p3, p4, p5, p6, ncol=2)

```

### This plot matrix brings all the combinations together in one easy to 
### view plot. Quality White Wines have an above average level of citric acid, 
### lower level of chlorides, higher level of alcohol, and medium to high level 
### of fixed acidity compared to the lower quality White Wines.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, alcohol, colour=quality_fac), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.95)))))

```

### Once again the Quality of White Wines but this time with the Density and 
### Alcohol switched around showing the strong linear relationship as well as the
### quality factor.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density, y = alcohol, color = quality_fac), data = WW) +
  geom_point(aes()) +
  scale_color_brewer(type = 'qual', palette = 2)

ggplot(aes(x = density, y = alcohol, colour = quality_fac), data = WW) +
  facet_wrap( ~ quality) +
  geom_point(aes()) +
  scale_color_brewer(type = 'qual', palette = 2)
```

### This is a better look at the different quality White Wines compared to   
### Density and Alcohol together and then seperate in order to see the 
### difference. I find it fascinating that alcohol and density are so closely 
### related to each other as well as the quality of white wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

alcohol.bucket = cut(WW$alcohol,
                        c(8, 9, 10.5, 12, 16))
ggplot(aes(x=factor(quality), y = citric.acid), data = WW) +
  geom_boxplot(aes(fill= alcohol.bucket))
  
alcohol.bucket2 = cut(WW$alcohol,
                     c(8, 9, 10.5, 12, 16))
ggplot(aes(x=factor(quality), y = density), data = WW) +
  geom_boxplot(aes(fill= alcohol.bucket2))

alcohol.bucket3 = cut(WW$alcohol,
                     c(8, 9, 10.5, 12, 16))
ggplot(aes(x=factor(quality), y = chlorides), data = WW) +
  geom_boxplot(aes(fill= alcohol.bucket3))
                     
```
### I was curious to look at the data in another way besides a scatterplot to
### makw sure my findings were correct.
### All three of these boxplots seems to back up my findings that higher citric
### acid, lower density (higher alcohol content), and lower chlorides make a 
### better quality White Wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, total.sulfur.dioxide, colour=alcohol), data = WW) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.95))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$total.sulfur.dioxide, 
                                           c(0.01,0.95)))))

```

### Density seems to be closely tied to the alcohol content as well as possibly
### total sulfur dioxide. The higher the alcohol content the less dense the 
### White Wines appear to be.


# Multivariate Analysis

## Talk about some of the relationships you observed in this part of the 
## investigation. Were there features that strengthened each other in terms of 
## looking at your feature(s) of interest?

### I was suprised to find that the higher quality White Wines seem to have a 
### higher alcohol content which in turn means a lower density. I thought that  
### the opposite would be true due to the taste of alcohol.
### I was also suprised to find that the higher quality White Wines had a medium 
### to high level of citric acid as well as low levels of chlorides(salt).

## Were there any interesting or surprising interactions between features?

### I thought it was definitely interesting that as the alcohol content goes 
### up the density goes down.

## OPTIONAL: Did you create any models with your dataset? Discuss the
## strengths and limitations of your model.

### I did not create a model.


# Final Plots and Summary

# Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

qplot(density, residual.sugar, data=WW) +
  geom_point(color='red', alpha = .5, size = 1) +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$residual.sugar, c(0.01,0.99))))) +
  geom_smooth(method='lm') +
  labs(x='Density (g/cm^3)', y='Residual Sugar (g/dm^3)', title='Strong Linear Relationship')
  
```

### The Density and Residual Sugar of the White Wines have a strong linear 
### relationship as shown in the above plot

# Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

ggplot(aes(x = density, y = alcohol, colour = quality_fac), data = WW) +
  geom_point(alpha = .5, size = 1.5, position = 'jitter') +
  geom_smooth(method='lm') +
  scale_x_continuous(lim=c(unname(quantile(WW$density, c(0.01,0.99))))) +
  scale_y_continuous(lim=c(unname(quantile(WW$alcohol, c(0.01,0.99))))) +
  scale_color_brewer(type = 'div', palette = 8) +
  labs(x='Density (g/cm^3)', y='Alcohol (%)', colour='Quality', 
       title='The Makeup of a Quality White Wine')

```

### The higher quality White Wines have a higher alcohol content and lower 
### density than the lower quality White Wines. It also shows a strong linear
### relationship between density and alcohol.

# Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

alcohol.bucket = cut(WW$alcohol,
                     c(8, 9, 10.5, 12, 16))
ggplot(aes(x=factor(quality), y = citric.acid), data = WW) +
  geom_boxplot(aes(fill= alcohol.bucket)) +
  labs(x='Quality', y='Citric Acid (g/dm^3)', title='Higher Citric Acid 
       Makes a Higher Quality White Wine', fill='Alcohol (%)')


alcohol.bucket3 = cut(WW$alcohol,
                      c(8, 9, 10.5, 12, 16))
ggplot(aes(x=factor(quality), y = chlorides), data = WW) +
  geom_boxplot(aes(fill= alcohol.bucket3)) +
  labs(x='Quality', y='Chlorides (g/dm^3)', fill='Alcohol (%)', title='Lower 
       Chlorides Makes a Higher Quality White Wine')

```

### Both of these boxplots seems to back up my findings that higher citric
### acid, higher alcohol content, and lower chlorides make a better quality
### White Wine.


# Reflection

### This dataset contained 4898 observations of White Wine Quality with 11
### inputs and 1 output. After exploring the data in detail I can say for certain 
### I know alot more about Wine than I have ever known. At first I was 
### concentrating strictly on what variables are needed to make a high quality
### wine then after more research I started wondering how the variables related
### with one another. I was very suprised to find that density is closely related
### to alcohol content as well as residual sugars. The more dense the wine was 
### the less alcohol content it contained.

### After examing the sulphates and sulfur dioxide I was very suprised to learn
### they are not closely correlated as it mentioned in the description of 
### attributes that sulphates can contribute to sulfur dioxide gas levels. It
### appears that density and total sulfur dioxide have a linear realtionship also
### that could be futhur examined.

### I had some trouble bringing together the multiple variables without including
### quality, but as I moved away from the only output variable (quality) it
### was apparent that there was some strong linear relationships between the
### other variables. I was also stumped as to why density, residual sugar, and
### alcohol are so closely related to each other until I started doing some 
### research and saw the linear relationships between the three. 

### I think there is opportunity for furthur understanding of the makeup of
### a good quality wine with more data on a wider number of white wines. Breaking 
### up the data into the 7 classes of whites would also allow you to gain more 
### understanding how the different variables that make up the wines react and
### come together to form a high quality wine. 
